Score-based tests of measurement invariance: Use in practice

### 2013-09-10 frontiers/frontiers.Rnw
update citations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@254 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

As currently implemented for SEM, the statistical tests described in
this paper can be applied to models that are estimated via a
multivariate normal or Wishart likelihood (or discrepancy) function,
with extension to other discrepancy functions being conceptually straightforward. 
The tests are carried out following model estimation, making use of output
associated with the fitted model.  In general, we fit a model that
restricts parameters to be equal across observations, then carry out a
posthoc test to examine whether specific parameters vary
across observations.  This procedure is similar in spirit to the
calculation of modification indices CITE and to Lagrange multiplier
tests CITE, and, in fact, those statistics can be viewed as  
special cases of the family described here.

Following model estimation, the tests primarily work on the {\em
  scores} of the fitted model; these are defined as
%% TODO indicate that this is evaluated at theta hat
\begin{equation}
\label{eq:score}
s(\hat{\theta};x_{i}) = \left ( \frac{\partial
      \ell(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial
      \ell(\theta;x_{i})}{\partial(\theta_{k})} \right )^{T},
\end{equation} 
where $\ell(\theta; x_i)$ is the likelihood associated with individual
$i$, $\theta$ is a $k$-dimensional parameter vector, and
$\hat{\theta}$ is the maximum likelihood estimate.  The cross-product
of these scores forms the ``meat'' for the calculation of robust
(Huber-White) standard errors. %TODO cite
\seeme{I don't understand this sentence. cross-product of each individual for each parameter? Why this 
is the standard error?}

@@

In this section, we demonstrate how the above tests can be carried out in~R, 
using the package lavaan \cite{lavaan11} for model estimation and strucchange 
\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
the applicability of adult gratitude scales to youth, available in the~R 
package psychotools CITE.  The data consist of responses to three adult 
gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
were specifically interested in whether the scales were measurement invariant 
across age.  Because the sample size at each age was unbalanced, the authors 
created age groups of approximately equal sample size.  In the examples below, 
we test for measurement invariance across these age groups.

>>>>

As currently implemented for SEM, the statistical tests described in
this paper can be applied to models that are estimated via a
multivariate normal or Wishart likelihood (or discrepancy) function,
with extension to other discrepancy functions being conceptually straightforward. 
The tests are carried out following model estimation, making use of output
associated with the fitted model.  In general, we fit a model that
restricts parameters to be equal across observations, then carry out a
posthoc test to examine whether specific parameters vary
across observations.  This procedure is similar in spirit to the
calculation of modification indices \cite{bentler90} and to Lagrange multiplier
tests \cite{sat89}, and, in fact, those statistics can be viewed as  
special cases of the family described here.

Following model estimation, the tests primarily work on the {\em
  scores} of the fitted model; these are defined as
%% TODO indicate that this is evaluated at theta hat
\begin{equation}
\label{eq:score}
s(\hat{\theta};x_{i}) = \left ( \frac{\partial
      \ell(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial
      \ell(\theta;x_{i})}{\partial(\theta_{k})} \right )^{T},
\end{equation} 
where $\ell(\theta; x_i)$ is the likelihood associated with individual
$i$, $\theta$ is a $k$-dimensional parameter vector, and
$\hat{\theta}$ is the maximum likelihood estimate.  The cross-product
of these scores forms the ``meat'' for the calculation of robust
(Huber-White) standard errors \cite{freed06}. \seeme{not sure about this citation}

@@

In this section, we demonstrate how the above tests can be carried out in~R, 
using the package lavaan \cite{lavaan11} for model estimation and strucchange 
\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
the applicability of adult gratitude scales to youth, available in the~R 
package psychotools \cite{zeilei11}.  The data consist of responses to three adult 
gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
were specifically interested in whether the scales were measurement invariant 
across age.  Because the sample size at each age was unbalanced, the authors 
created age groups of approximately equal sample size.  In the examples below, 
we test for measurement invariance across these age groups.
